\chapter*{Acknowledgements}
\label{acknowledgements}

Firstly I would like to thank my two main supervisors, Kyle Dexter and Casey Ryan, for guiding me through my PhD and along the road to becoming a real academic. My PhD experience has been enriched by the many opportunities they sent my way, and by the freedom I was given to pursue my own interests throughout the project. With their guidance I have learned that research is about collaboration, creativity, and exploration. 

For support during fieldwork I thank Francisco Maiato Gon\c{c}alves, who looked after me during my two visits to Angola, and who made me feel at home with his family and friends. I also thank Jos\'{e} Tchamba, Abel Cahali, Fortunato da Silva, and Henriques Jacinto for their support and friendship during our most difficult field campaign. For braving the bees, snakes, and rain, while measuring many thousands of trees in Bicuar National Park. I also thank the rangers and support staff at Bicuar National Park for guiding us and always providing delicious food at the end of the day. 

For support during my fieldwork in Kilwa District, Tanzania I thank all those who helped me from Mpingo Conservation \& Development Initiative in Kilwa Masoko. In particular, I extend my deepest gratitude to Paula Whyte for helping me navigate the impenetrable Tanzanian bureaucracy, and for being a friend during the long days at the Slow Leopard as I waited for permits. For help during the fieldwork itself I thank Emanuel Mlimitho for his impressive work ethic and attention to detail, and Abdala Mpotola, village chairman of Nainokwe, for his help with identifying tree species and digging us out of many situations on the road.

Thank you to the NERC Geophysical Equipment Facility for their help with the terrestrial LiDAR, and for trusting me with such a delicate and expensive piece of equipment. Thank you also to Steven Hancock who provided so much help in analysing the LiDAR data, without which I could not have attempted such a huge task.

This PhD was funded primarily by the NERC E3 Doctoral Training Partnership. Thank you to the Davis Expedition Fund which provided funding for the Tanzania portion of the fieldwork, and to the Centenary Agroforestry 89 Fund for the Angola part of the fieldwork. Thank you to National Geographic who provided funding for the creation of the Bicuar National Park permanent plots, which will hopefully benefit many other researchers in years to come, and without which a large portion of this thesis would not have come to fruition.

Finally, thank you to Rosie for your unwavering support, for talking me through re-wiring a generator from 5000 miles away, for listening with apparent interest to many hours of chatter about LiDAR and canopy architecture, and for being my partner as we both navigate our way through our PhD programmes. During the final months of writing up you kept me healthy and positive. Now it is my turn to do the same for you.
